---
output: html_document
editor_options: 
  chunk_output_type: console
---

# BIOE215 Week 3 Assessment

*Aspen Ellis*

```{r}
library(tidyverse)
library(here) #this is how to get filepaths to play nice in projects
bor_nestlings <- read_csv(here::here("data/73_species.csv"))
bor_traits <- read.csv(here::here("data/Traits_73_species.csv"))
avonet <- readxl::read_excel(here::here("data/AVONET_20Supplementary_20dataset_201.xlsx"), sheet = "AVONET1_BirdLife")


```

*Are nestlings showing up earlier in the year over time?*

```{r}
all_birds_trend <- bor_nestlings %>% 
  group_by(Year) %>% 
  summarize(mean_doy = mean(Dayofyear))

ggplot(all_birds_trend, aes(Year, mean_doy)) + 
  geom_point() +
  geom_smooth(method = "lm")
```

Yes they are!

But that was grouped, why don't we look by species:

```{r}
species_trends <- bor_nestlings %>% 
  group_by(Year, Species) %>% 
  summarize(mean_doy = mean(Dayofyear),
            .groups = "drop")

ggplot(species_trends, aes(Year, mean_doy, color = Species)) + 
  geom_point() +
  geom_smooth(method = "lm")
```

yeaaarrrgh!! chaos. let's just look at the five spp that we have the most data for:

```{r}
data_richness <- bor_nestlings %>% 
  count(Species)

most_rich <- data_richness %>% 
  arrange(desc(n)) %>% 
  slice(1:5)

most_rich_trends <- bor_nestlings %>% 
  filter(Species %in% most_rich$Species) %>% 
  group_by(Species, Year) %>% 
  summarize(mean_doy = mean(Dayofyear), 
            .groups = "drop")

ggplot(most_rich_trends, aes(Year, mean_doy, color = Species)) +
  geom_point() +
  geom_smooth(method = "lm")
```

much nicer :) most are getting earlier but LARARG is getting later. What species have the strongest trends?

```{r}
# I'm giving you a *function* to help here. You'll learn more about those next week.

# Finds the slope of the relationship between y and x
trend <- function(x, y) {
  xy_lm <- lm(y ~ x)
  coef(xy_lm)[2]
}

# Calculate the trend for all species
bor_trends <- species_trends %>% 
  group_by(Species) %>% 
  summarize(doy_trend = trend(Year, mean_doy))

```

Spot check two species

```{r}
soi <- c("ARDCIN", "LARMIN")
species_trends %>% 
  filter(Species %in% soi) %>% 
  ggplot(aes(Year, mean_doy, color = Species)) +
  geom_point() +
  geom_smooth(method = "lm")
```

My turn apparently: finding the species with most positive and negative trends and just plot those

```{r}
nrow_bor_trends <- nrow(bor_trends) # Use this later
bor_extreme <- bor_trends %>% 
  # Sort by the day of year trend
  arrange(doy_trend) %>% 
  # Keep just the first (most negative trend) and last (most positive trend) rows
  slice(c(1, nrow_bor_trends))

# Now plot them
species_trends %>% 
  filter(Species %in% bor_extreme$Species) %>% 
  ggplot(aes(Year, mean_doy, color = Species)) + 
    geom_point() +
    geom_smooth(method = "lm")
```

Next we're going to test the hypothesis that smaller birds have more flexible phenology, i.e. the absolute value of the trend is correlated with smaller body sizes.

To test our hypothesis, we need to add AVONET data to our phenology data by joining. The keys are a bit tricky here!

`bor_nestlings` has a column called `Species` with a 6-letter code.

`bor_traits` has a column called `Abbreviation` with the 6-letter code and a column called `` `Scientific name` `` with the binomial name.

`avonet` has column called `Species1` with the binomial name.

We need to join `bor_nestlings` to `bor_traits`, then join with `avonet`.

```{r}
# First, select and rename important columns
avonet_size <- select(avonet, SciName = Species1, Mass_g = Mass)
bor_sciname <- select(bor_traits, 
                      Species = Abbreviation, 
                      SciName = Scientific.name)
```

Now join them all together.

```{r}
bor_trends_size <- bor_trends %>% 
  left_join(bor_sciname, bor_nestlings, by = "Species") %>% 
  left_join(avonet_size, by = "SciName") %>% 
  mutate(abs_trend = abs(doy_trend))

# Plot it
ggplot(bor_trends_size, aes(Mass_g, abs_trend)) +
  geom_point() +
  geom_smooth(method = "lm")
```

Hard to interpret - make a new variable, `bor_trends_size2`, that removes the missing values in `Mass_g` and keeps only birds smaller than 2000 g with an absolute trend less than 1.

```{r}
bor_trends_size2 <- bor_trends_size %>% 
  filter(!is.na(Mass_g)) %>% 
  filter(Mass_g<2000) %>% 
  filter(abs_trend<1)
  
ggplot(bor_trends_size2, aes(Mass_g, abs_trend)) +
  geom_point() +
  geom_smooth(method = "lm")
```

Still kinda hard to say lol I'd say it doesn't look like very compelling support for our hypothesis.

OK TTFN!
